Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-04 journal paper/content/sec-intro.tex
Full draft up to and including "theory results".

The main high-level question that we focus on in this paper is: 
\begin{align*}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
\end{align*}

@@

\subsection{Our model}
\label{sec:intro-model}

@@

\item Principals simultaneously announce their bandit algorithms before round $1$, and cannot change them afterwards.
    Each principal only observes agents that chose this principal.

\item Agents do not directly observe other agents' rewards, but may have access to each platforms' reputation score (more on this in    Section~\ref{sec:intro-discussion}).
\end{itemize}

\xhdr{Technology: multi-armed bandit algorithms.}
While some bandit algorithms are better than others, these comparisons are rather subtle. Indeed, some algorithms may be better for some problem instances and/or time intervals, and worse for some others.

@@

\noindent Greedy algorithms are terrible for a wide variety of problem instances, precisely because they never explore. Adaptive-exploration algorithms are optimal in the worst case, and exponentially improve for ``easy" problem instances. Exploration-separated algorithms learn at a reasonable but mediocre rate across all problem instances, and fail to improve for the ``easy" ones.

\ascomment{stopped here.}

@@

@@

\xhdr{Model variants.}
We consider two model variants, which determine agents' decision rule and their information sets: 
\ascomment{copied-and-pasted, will edit later.}

\begin{description}
\item[\TheoryModel] 
There is a common Bayesian prior on the reward distributions.
Agents do not receive any other information and choose between the principals using their knowledge of $t$ and the principals' algorithms.
\item[\ExptsModel] agents have access to a reputation score for each principal, which is a sliding window average of the rewards experienced by previous agents that have visited this principal.
\end{description}

We consider two separate variants of the model in order to provide a more thorough investigation of the tension between exploration and competition. The expectation choice model is tractable enough to allow us to obtain theoretical results with ``asymptotic" flavor. However, for the sake of analytical tractability, we make the unrealistic simplification that users do not observe any signals about firms' ongoing performance and it is difficult to analyze important economic mechanisms such as asymmetries in the timing of entry.

>>>>

The main high-level question that we focus on in this paper is: 
\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
\end{align}

@@

\subsection{Our model: competing bandits}
\label{sec:intro-model}

@@

\item Principals simultaneously announce their bandit algorithms before round $1$, and cannot change them afterwards.
    Each principal only observes agents that chose this principal. Agents have little or no information about other agents' choices and rewards.
%, but may have access to each platforms' reputation score (more on this in    Section~\ref{sec:intro-discussion}).
\end{itemize}

\noindent We investigate several model variants within this framework, where we vary agents' decision rule and/or give a first-mover advantage to one of the principals.

%We consider two model variants, which determine agents' decision rule and their information sets. In \emTheoryModel, there is a common Bayesian prior on the reward distributions. Agents do not receive any other information and choose between the principals using their knowledge of $t$ and the principals' algorithms. In \emExptsModel, agents have access to a reputation score for each principal, which is a sliding window average of the rewards experienced by previous agents that have visited this principal. The former variant used for the theoretical results, the latter for simulations.

\xhdr{Technology: multi-armed bandit algorithms.}
To flesh out \eqref{eq:main-Q}, let us elaborate what we mean by `better' bandit algorithms. In general, comparisons between bandit algorithms are rather subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others.

@@

\noindent Greedy algorithms are terrible for a wide variety of problem instances, precisely because they never explore. Exploration-separated algorithms learn at a reasonable but mediocre rate across all problem instances. Adaptive-exploration algorithms are optimal in the worst case, and exponentially improve for ``easy" problem instances.

@@

\xhdr{Theoretical results.} 
We endow agents with Bayesian rationality, arguably the most natural modeling approach for a theoretical investigation. We consider a `bare-bones' variant of Bayesian rationality: agents have a common Bayesian prior on reward distributions, and know the principals' algorithms. For simplicity, agents do not receive any information about previous agents' choices and rewards. Each agent knows the round (s)he arrives in, and computes principals' Bayesian-expected rewards for this round.

Our results depend crucially on agents' decision rule to choose among the two principals. The most obvious decision rule maximizes the Bayesian-expected reward; we refer to it as \HardMax. We find that \HardMax is not conducive to adopting better algorithms: each principal's dominant strategy is to choose the greedy algorithm. Further, we show that \HardMax is very sensitive to tie-breaking: if the tie-breaking is probabilistically biased in favor of one principal, then this principal has a simple ``winning strategy" no matter what the other principal does.

Next, we dilute the \HardMax agents with a small fraction of ``random agents" who choose a principal uniformly at random. We call this model \HardMaxRandom. We find that better algorithms help in a big way: a sufficiently better algorithm is guaranteed to win all non-random agents after an initial learning phase. However, there is a substantial caveat: one can defeat any algorithm by interleaving it with the greedy algorithm. This has two undesirable corollaries: a better algorithm may sometimes lose in competition, and a pure Nash equilibrium typically does not exist.

We further relax the decision rule so that the probability of choosing a given principal varies smoothly as a function of the difference between  principals' Bayesian-expected rewards; we call it \SoftMaxRandom. For this model, the ``better algorithm wins" result holds under much weaker assumptions on what constitutes a better algorithm. This is the most technical result of the paper. The competition in this setting is necessarily much more relaxed: typically, both principals attract approximately half of the agents as time goes by (but a better algorithm would attract slightly more).

All results extend to a much more general version of the multi-armed bandit problem in which the principal may observe additional feedback before and/or after each decision, as long as the feedback distribution does not change over time. In most results, principal's utility may depend on both the market share and agents' rewards.

\xhdr{Economic interpretation: the inverted-U relationship.}
Interpreting the adoption of better algorithms as ``innovation", our findings can be framed in terms of the inverted-U relationship between competition and innovation, as in Figure~\ref{fig:inverted-U}.%
\footnote{Here, ``innovation" refers to adoption of a better technology at a substantial R\&D expense to a given firm. It is not salient whether similar ideas and/or technologies already exist elsewhere. Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms themselves are well-known in the research literature \citep[\eg see][]{DS-arxiv}.}
This is a well-established concept in the economics literature, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better.

Our decision rules differ in terms of rationality: from fully rational decisions with \HardMax to relaxed rationality with \HardMaxRandom to an even more relaxed rationality with \SoftMaxRandom. The same distinctions also control the severity of competition between the principals: from cut-throat competition with \HardMax to a more relaxed competition with \HardMaxRandom, to an even more relaxed competition with \SoftMaxRandom. Indeed, with \HardMax you lose all customers as soon as you fall behind in performance, with \HardMaxRandom you get some small market share no matter what, and with \SoftMaxRandom you are further guaranteed a market share close to $\tfrac12$ as long as your performance is not much worse than the competition. The uniform choice among principals corresponds to no rationality and no competition.

We identify the inverted-U relationship driven by the rationality/competitiveness distinctions outlined above: from \HardMax to \HardMaxRandom to \SoftMaxRandom to \Uniform. We also find another, technically different inverted-U relationship which zeroes in on the \HardMaxRandom model. We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.

These inverted-U relationships are driven by different aspects in our model than the ones in the existing literature in economics. The latter focuses on the tradeoff between the R\&D costs and the benefits that the improved technology provides in the competition. In our case, the barriers for innovations arise entirely from the reputational consequences of exploration in competition, even in the absence of R\&D costs.

\asedit{Stopped here.}

@@

\xhdr{Model variants.}
We consider two separate variants of the model in order to provide a more thorough investigation of the tension between exploration and competition. The expectation choice model is tractable enough to allow us to obtain theoretical results with ``asymptotic" flavor. However, for the sake of analytical tractability, we make the unrealistic simplification that users do not observe any signals about firms' ongoing performance and it is difficult to analyze important economic mechanisms such as asymmetries in the timing of entry.
